One-Particle Density Matrix Polarization Susceptibility Tensors

### 2018-06-03 papers/1_dmatpol/paper-r1.tex
Revision

\documentclass[aip,amsmath,amssymb,reprint]{revtex4-1}
%\documentclass[aip,reprint]{revtex4-1}

@@

Molecular\hyp{}level exploration of condensed phase phenomena becomes increasingly unavoidable
as technology and medicine continue to develop. Unfortunately, the computational cost of accurate
electronic structure calculations grows extremely fast with the number of particles, greatly limiting
the use of the state\hyp{}of\hyp{}the\hyp{}art first\hyp{}principles theories 
when large and complex chemical systems such as
solutions and biomolecules come into play. Although the advent in the computer technology 
brought some hopes, it is still insufficient to follow up the expensive needs.
Due to this reason, it has been of particular importance nowadays to develop bridging theoretical techniques that link
extremely accurate but localized pieces of information with the heuristic knowledge of an atomistic organisation 
on a relatively large spatial scales, typically ranging beyond few nanometers.

A class of molecular dynamics simulation techniques was developed over the years. That was possible
due to the fruitful incorporation of accurate \emph{ab initio} methods and a 
semi\hyp{}classical notion of molecular motion combined with extensive experimental and semi\hyp{}empirical
refinings. The advent of hybrid methods such as quantum mechanical\hyp{}molecular mechanical (QM/MM)
technique was a revolution in computational chemistry and allowed insightful theoretical studies
of quantum mechanical effects in biological systems because the high\hyp{}level 
and costly quantum mechanical calculations
could be combined with low\hyp{}level and cheap classical dynamics. 
The size of quantum level can be extended without compromising accuracy and maintaining good performance
by using vast family of fragmentation methods, such as effective fragment potentials (EFP).
The most difficult to tackle part of Hamiltonian can also be embedded within a widely understood effective environment
leading to various embedding methodologies, realising similar goals.

However, even with the modern advances in computational chemistry, 
%accurate description of environmental effects
accurate description of phenomena due to complex heterogeneous molecular environments 
remains a challenge in the following areas: (i) describing the electronically excited states 
(electronic solvatochromism), (ii) understanding
the vibrational properties of molecules in ground and electronically excited states
(vibrational or vibronic solvatochromisms). EFP methods are limited to the ground\hyp{}state
chemistry at the Hartree\hyp{}Fock (HF) or density functional theory (DFT) levels.
It is also quite difficult to generalize the embedding potential operator
because of complicated nature of intermolecular interactions, that theoretically
cannot be solely envisioned as a sum of electrostatic, dispersion and Pauli effects.
Therefore, Hamiltonian embedding is often realised by Coulombic embedding with polarization
of QM level, and only rarely the Pauli exclusion principle is taken into account. Although
such approach is very accurate in certain cases, it is not guaranteed that it is sufficient
for all cases.

In this Work, we explore the susceptibility of the molecule's electronic density distribution
to polarize due to external electrostatic perturbation from first principles. 
The deformation density is expressed in terms
of the density matrix susceptibility (DMS) tensors that are properties of isolated molecules and can be
extracted from quantum chemistry calculations. The focus on the density (matrix) instead of the energy
(scalar) is beneficial since it provides the directional tensor description
of the polarization that could be combined with other effects such as Pauli repulsion or dispersion,
in contrast to the standard interaction energy approaches. We show that accurate polarization 
of the one\hyp{}particle density
matrix is achievable and can be realized as long as the one\hyp{}particle density matrices
are available. It is believed that the proposed method
could be a basis for a construction of an 
alternative approach in fragment\hyp{}based
modelling of extended molecular aggregates that is not limitted to the ground state chemistry.
We introduce the necessary theory and derive the \emph{ab initio} density matrix susceptibility model
at the HF approximation in Section II. This model serves as a basis for generalization to the 
arbitrary one\hyp{}electron density with electronic correlation which is derived in Section II. 
The implementation details in a computer code as well as the validation procedures are detailed
in Section III. We then discuss the performance of the models in Section IV based on a few small molecules
in constant and varying electric field distributions. Finally, we conclude the presentation wit a few 
remarks for a future development in Section V.

@@

In the Hartree\hyp{}Fock theory, ${\bf D}$ is approximated as
%
\begin{equation}
 {\bf D} = {\bf C}{\bf C}^\dagger \;,
\end{equation}
%
where ${\bf C}$ describes the occupied molecular orbitals as linear combinations
of atomic orbitals (AO's), referred here to as the LCAO-MO matrix.
${\bf D}$ is an idempotent matrix and if expressed in orthogonal basis the following holds
%
\begin{equation}
 {\bf D}^2 = {\bf D} \;.
\end{equation}
%
The change in the molecular orbitals can be parameterized as suggested by McWeeny\cite{McWeeny.RevModPhys.1960}
by defining the idempotency\hyp{}preserving variation
%
\begin{equation}
 \delta {\bf C} = {\BM\Delta} \cdot {\bf C}^{(0)} \;,
\end{equation}
%
where $\BM\Delta$ is a square non\hyp{}singular and non\hyp{}symmetric matrix.
McWeeny has shown that
%
\begin{equation} \label{e:dmat-change.exact}
 \delta {\bf D} = -{\bf D}^{(0)} + \left[ {\bf D}^{(0)} + {\bf v} \right]
                                   \left[ {\bf 1} + {\bf v}^\dagger{\bf v} \right]^{-1}
                                   \left[ {\bf D}^{(0)} + {\bf v}^\dagger \right] \;,
\end{equation}
%
where the auxiliary vector ${\bf v}$ is defined as
%
\begin{equation}
 {\bf v} \equiv \left[ {\bf 1} - {\bf D}^{(0)} \right] {\BM\Delta} {\bf D}^{(0)}  \;.
\end{equation}
%
Expanding Eq.~\eqref{e:dmat-change.exact} in a Taylor series around ${\bf v}={\bf 0}$ and
truncating it at linear terms with respect to ${\BM\Delta}$ gives
%
\begin{subequations} 
 \begin{align}
 \delta {\bf D} &\cong \left[ {\bf 1} - {\bf D}^{(0)} \right] {\BM\Delta} {\bf D}^{(0)} + 
                        {\bf D}^{(0)} {\BM\Delta}^\dagger \left[ {\bf 1} - {\bf D}^{(0)} \right]  
 \label{e:dD-1} \\  &= 
  \delta {\bf C}  {\bf C}^{(0)\dagger} + {\bf C}^{(0)}\delta {\bf C}^\dagger
           - {\bf D}^{(0)} \delta {\bf C}  {\bf C}^{{(0)}\dagger} - {\bf C}^{(0)}\delta {\bf C}^\dagger {\bf D}^{(0)} 
 \label{e:dD-2}
 \end{align}
\end{subequations}
%
The latter equality is true only if the basis set functions are orthogonal, i.e., 
${\bf C}^{(0)\dagger}{\bf C}^{(0)}={\bf 1}$,
which will be the case throughout the entire work, unless stated differently.
Substituting the result from Eq.~\eqref{e:dD-1}
into Eq.~\eqref{e:dmu-exact} leads to
%%
%\begin{equation} \label{e:delta-Taylor}
% \delta {\bf D} = \left[ {\bf v} + {\bf v}^\dagger \right]
%                + \left[ {\bf v}{\bf v}^\dagger - {\bf v}^\dagger{\bf v}\right] + \ldots 
%\end{equation}
%
\begin{multline} \label{e:dmu-4-exact.linear-approximation}
 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong
   \Tr{ 
    \left[ 
         {\mathbb{M}} {\BM\Delta} {\bf D}^{(0)}  
    \right] }
%
  +\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta}^{\dagger}
    \right] } \\
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta} {\bf D}^{(0)}
    \right] }
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta}^{\dagger} {\bf D}^{(0)}
    \right] } \;,
\end{multline}
%
It can be proven (see Appendix~\ref{a:orig-dep}) that such obtained
induced dipole moment is independent on the origin. 
This quarantees that the theory fulfills the total charge conservation requirement.
Therefore,
in the above result and also in later course of the Paper the dipole origin will be omitted. 
%where the dipole origin was omitted for notational clarity.

@@

      = \left[  {\bf C}^\dagger \mathbb{M} \left[ \bf{1} - \bf{D}^{(0)} \right] \right]_{i\alpha}

@@

The theory that is necessary to validate the density matrix susceptibility models from Section~\ref{s:2}
was implemented in our in\hyp{}house plugin to {\sc Psi4} quantum chemistry program.
Water molecule, energy\hyp{}optimized by using the HF/6-311++G** method, 
was selected as a test system which was subject to 
electrostatic perturbations due to random collections of point charges
generating non\hyp{}uniform electric fields. 
Cartesian coordinates of perturbing charges were drawn from uniform distribution of a set of fourty
three\hyp{}dimensional Euclidean points within a sphere of radius 6.0 Bohrs
and origin in the center of mass of water molecule.
In selection of random points, space inside the union of spheres surrounding
oxygen and hydrogen atoms with radii of 4.5 and 4.0 Bohrs, respectively, was
kept free from any point charges due to potential problems with the convergence of the 
self\hyp{}consistent field (SCF) calculations.
Charge values were selected from the one\hyp{}dimensional uniform distribution
of double precision numbers in the interval of $\vert q_{\rm max} \vert$, where $q_{\rm max}$
was tuned to obtain a desired ensemble of weak, moderate and strong electric field magnitudes
(with $q_{\rm max}=0.05$, $0.2$ and $0.4$ a.u., accordingly; cf. Table~\ref{t:fields}).

%For finding a general unitary transformations,
%100 random point charge distributions were selected for which an exact density matrices
%were calculated.
In order to validate all the models discussed in this Contribution,
the electric field\hyp{}induced interaction energy, 
the dipole moment and the quadrupole moment were analysed in detail.
The electric field\hyp{}induced interaction energy at the HF level, $\delta E$,
was computed from the following exact formula:
%
\begin{multline}\label{e:dE}
 \delta E[{\bf F}] = E_{\rm nuc}({\bf F}) +
                     \Tr{\left\{ \delta {\bf D} \cdot
                                \left( {\bf H}^{(0)} + {\bf f}^{(0)} \right) \right\}} \\
                   + \Tr{\left\{ \left( {\bf D}^{(0)} + \delta {\bf D} \right) \cdot
                                \left( 2{\bf V} + 2\delta {\bf J} - \delta {\bf K}\right) \right\}} \;.
\end{multline}
%
In Eq.~\eqref{e:dE}, ${\bf f}^{(0)}$ and ${\bf H}^{(0)}$ are the Fock matrix and 
the one\hyp{}electron Hamiltonian matrix of an unperturbed HF ground state,
${\bf V}$ is the one\hyp{}electron external electrostatic potential matrix whereas $\delta {\bf J}$
and $\delta {\bf K}$ are electric field\hyp{}induced changes in the Coulomb and the exchange HF matrices due to 
the polarization of the electron density distribution, respectively. $E_{\rm nuc}({\bf F})$ denotes the interaction energy
due to the nuclei.
In order to compare the performance of the distributed polarizability approach,
the approximate interaction energy was also evaluated by using the expression:
%
\begin{multline}\label{e:dE-dpol}
 \delta E[{\bf F}] \approx E_{\rm nuc}({\bf F}) + 
                     2 \Tr{\left\{ {\bf D}^{(0)} {\bf V}\right\} }  \\
                   - \frac{1}{2} \sum_i^{\rm occ} {\BM \alpha}_i : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_i) \;,
\end{multline}
%
in which ${\BM \alpha}_i$ is the distributed anisotropic polarizability tensor associated with the
$i$th localized molecular orbital (LMO) centroid (see Eq.~\eqref{e:dmu-dpol}). 
%, ${\bf r}_i = \tbraket{\phi_i}{\bf r}{\phi_i}$.
The distributed polarizabilities were computed by utilizing the coupled\hyp{}perturbed Hartree\hyp{}Fock
method. Orbital localization was performed by using the Boys method. Note that in Eq.~\eqref{e:dE-dpol}, 
multipole approximation is introduced only for the part associated with the electric field\hyp{}induced
change in the one\hyp{}particle density matrix, i.e., the polarization of the electron density distribution.
%This is because we focus in this Work on the pure polarization process. Therefore, comparison between
%the results obtained by using Eq.~\eqref{e:dE-dpol} and Eq.~\eqref{e:dE} will reveal

The electric field\hyp{}induced dipole and primitive quadrupole moments were evaluated from the following expressions:
%
\begin{subequations}\label{e:dmult}
  \begin{align}
   \delta \upmu_{u}   &= 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \tbraket{\alpha}{u}{\beta}  \;,\\
   \delta \Theta_{uw} &= 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \tbraket{\alpha}{(u-u_{\rm CM})(w-w_{\rm CM})}{\beta} \;. 
  \end{align}
\end{subequations}
%
where the subscript CM denotes the center of mass of water molecule.
When the distributed polarizabilities were used instead of the polarization\hyp{}induced 
difference density matrices, primitive induced quadrupole moments were computed from
the induced dipole moments according to
%
\begin{equation}
  \delta {\BM \Theta} = \sum_i^{\rm occ} 
   \left\{ 
       \delta {\BM \upmu}_i \otimes \left( {\bf r}_i - {\bf r}_{\rm CM} \right) 
      + \left( {\bf r}_i - {\bf r}_{\rm CM} \right) \otimes \delta {\BM \upmu}_i
   \right\} \;.
\end{equation}
%
%where ${\bf r}_{\rm CM}$ is the center of mass of water molecule.

Primitive quadrupole moments were subsequently transformed to their traceless forms by
%
\begin{equation}
  \delta \widetilde{\Theta}_{uw} = \delta \Theta_{uw} - \frac{1}{3} \delta_{uw} \Tr{\BM\Theta} \;.
\end{equation}
%
To compute the interaction energy and the induced multipole moments, either exact or approximated forms
of the difference density matrix from Eq.~\eqref{e:final-model.General}
were used for a particular electric field distribution. The statistical evaluation of
errors with respect to the benchmark (full HF calculations) were based on 
the following statistical quality descriptors:
%
\begin{enumerate}
 \item Root Mean Square Energy Error (RMSE)
   \begin{equation}\label{e:rmse}
     {\rm RMSE} = \sqrt{ \frac{1}{N_{\rm max}} \sum_N \left( \delta E^{(N)} - \overline{\delta E^{(N)}}\right)^2 }
   \end{equation}
 \item Root Mean Square Dipole Error (RMSD)
   \begin{equation}\label{e:rmsd}
     {\rm RMSD} = \sqrt{ \frac{1}{N_{\rm max}}\sum_N \lvert \delta {\BM\upmu}^{(N)} - \overline{\delta {\BM\upmu}^{(N)}}\rvert^2 }
   \end{equation}
 \item Root Mean Square Quadrupole Error (RMSQ)
   \begin{equation}\label{e:rmsq}
     {\rm RMSQ} = \sqrt{ \frac{1}{N_{\rm max}}\sum_N \lvert \delta \widetilde{\BM\Theta}^{(N)} - \overline{\delta \widetilde{\BM\Theta}^{(N)}}\rvert^2 }
   \end{equation}
\end{enumerate}
%
The first describes the ability of the model to reproduce the exact energetics of the system, 
wherease the latter two contain crucial information on how well the deformation of the one\hyp{}electron
density is described. For Eqs.~\eqref{e:rmsd} and \eqref{e:rmsq} the absolute magnitudes
are defined as
%
\begin{subequations}\label{e:absmagn}
  \begin{align}
   \lvert {\BM\upmu} \rvert &= \sqrt{\mu_x^2+\mu_y^2+\mu_z^2} \;,\\
   \lvert {\widetilde{\BM\Theta}}\rvert &= 
          \sqrt{\widetilde{\Theta}_{zz}^2 + \frac{1}{3}\left(\widetilde{\Theta}_{xx}-\widetilde{\Theta}_{yy}\right)^2 
          + \frac{4}{3}\left( \widetilde{\Theta}_{xy}^2 + \widetilde{\Theta}_{xz}^2 + \widetilde{\Theta}_{yz}^2 \right)} \;. 
  \end{align}
\end{subequations}
%
Throughout the Work, the Hartree\hyp{}Fock level was assumed and the 6-311++G** basis set
was used. For all the calculations of the density matrix polarization susceptibility tensors
from Eqs.~\eqref{e:final-model.HF} and \eqref{e:susceptibility-B}, 
atomic basis was orthogonalized by using the L{\"o}wdin symmetric orthogonalization scheme.

>>>>

\documentclass[aip,amsmath,amssymb,reprint,floatfix]{revtex4-1}
%\documentclass[aip,reprint]{revtex4-1}

@@

Molecular\hyp{}level exploration of condensed phase phenomena becomes increasingly unavoidable
as technology and medicine continue to develop. Unfortunately, the computational cost of accurate
electronic structure calculations grows extremely fast with the number of particles, greatly limiting
the use of the state\hyp{}of\hyp{}the\hyp{}art first\hyp{}principles theories 
when large and complex chemical systems such as
solutions and biomolecules come into play. Although the advent in the computer technology 
brought some hopes, it is still insufficient to follow up the expensive needs
of computational chemistry.
Due to this reason, it has been of particular importance nowadays 
to develop bridging theoretical techniques that link
extremely accurate but localized pieces of information 
with the heuristic knowledge of an atomistic organisation and intermolecular interaction
on a relatively large spatial scales, typically ranging beyond few nanometers.

The advent of hybrid methods such as quantum mechanical\hyp{}molecular mechanical (QM/MM)
technique was a revolution in computational chemistry and allowed insightful theoretical studies
of quantum mechanical effects in biological systems because the high\hyp{}level
and costly quantum mechanical calculations
could be combined with low\hyp{}level and cheap classical dynamics.
That was possible due to the fruitful incorporation of accurate \emph{ab initio} methods and a 
semi\hyp{}classical notion of molecular motion combined with extensive experimental and semi\hyp{}empirical
refinings.
Understanding physics that rules intermolecular interactions helped to develop
a wide family of force fields, and served a basis for embedding Hamiltonian
within an effective environment. 
%The size of quantum level can be extended without compromising accuracy and maintaining good performance
%by using vast family of fragmentation methods, such as effective fragment potentials (EFP).
%The most difficult to tackle part of Hamiltonian can also be embedded within a widely understood effective environment
%leading to various embedding methodologies, realising similar goals.

However, even with the modern advances in computational chemistry, 
%accurate description of environmental effects
accurate description of phenomena due to complex heterogeneous molecular environments 
remains a challenge in the following areas: (i) describing the electronically excited states 
(electronic solvatochromism), (ii) understanding
the vibrational properties of molecules in ground and electronically excited states
(vibrational or vibronic solvatochromisms). 
Interface between QM and MM levels is usually realized by electrostatic embedding
and only rarely includes the Pauli exclusion principle and dispersion. 
Unfortunately, it is quite difficult to generalize the embedding potential operator
because of complicated nature of intermolecular interactions, that theoretically
cannot be solely envisioned as a sum of electrostatic, dispersion and Pauli effects.
Perhaps the most frequently used efficient and accurate \emph{ab initio} fragmentation methods,
the effective fragment potential (EFP) methods, are limited to the ground\hyp{}state
chemistry at the Hartree\hyp{}Fock (HF) or density functional theory (DFT) levels,
and, while they can approximately describe the environment composed of molecules in their ground states,
they cannot be used for highly correlated wavefunctions such as electronically excited states.
%Therefore, Hamiltonian embedding is often realised by Coulombic embedding with polarization
%of QM level, and only rarely the Pauli exclusion principle is taken into account. Although
%such approach is very accurate in certain cases, it is not guaranteed that it is sufficient
%for all cases.

Most of efforts to theoretically treat extended molecular aggregates are rooted in 
energy\hyp{}based philosophy. However,
in the past decade there has been considerable interest in density\hyp{}based approaches
of a wide scope. In particular, the importance of
one\hyp{}electron densities in chemistry is well known and acknowledged in the community,
though still not solved exactly.\cite{Holas.March.PhysRevA.1991}
%that seem to have an important advantage over the standard energetic approach.
This Work challenges in establishing a connection bewteen the one\hyp{}particle
density and the perturbation in the environment through introducing the
density matrix susceptibility (DMS) tensors.
The associated change in the one\hyp{}particle density matrix, $\delta {\bf D}$, is postulated to be
deconvoluted into a product form of the perturbation susceptibility ${\bf B}_i$
and the environment\hyp{}induced perturbation on a molecule, $\delta {\bf P}_i$.
%and induced dipole moments $\delta {\BM\upmu}_i$ due to electric fields ${\bf E}_i$ 
%at the centroids of the occupied molecular orbitals,
%
\begin{equation*} \label{e:initial-model}
 \delta {\bf D} \approx \sum_i^{\text{Sites}} \text{[DMS of the $i$th site]} 
  \cdot  \delta {\bf P}_i
  %\rightarrow \sum_i^{\rm occ} {\bf B}^{(i)}_{\mu\nu} \cdot {\bf E}_i \;.
\end{equation*}
%
While $\delta {\bf P}_i$ contain critical information 
about the actual distortion of the electronic cloud,
perturbation susceptibility is assumed to be a property of unperturbed 
molecule and could be precomputed only once. 
This could be adapted into an efficient effective 
one\hyp{}electron fragment potential, in principle applicable to an arbitrary
level of theory, as long as the one\hyp{}particle densities are available.
%
%a simple first\hyp{}principles model for prediction 
%of the perturbation\hyp{}induced one\hyp{}electron density
%change of structurally unrelaxed molecule is proposed here.

The simplest, yet perhaps the most relevant for chemistry, type of perturbation 
is electrostatic perturbation.
It is worth 
mentioning that, at that moment, there is no simple \emph{ab initio}
method of computation of electron deformation densities without considering the entire molecular 
aggregate.\cite{Horn.Yuezhi.Head-Gordon.JCP.2016,Horn.Head-Gordon.JCP.2015}
In this Work, we explore the susceptibility of the molecule's electronic density distribution
to polarize due to external electrostatic perturbation from first principles. 
The deformation density is expressed in terms
of the one\hyp{}particle density matrix polarization susceptibility tensors 
that are properties of isolated molecules and can be
extracted from quantum chemistry calculations. The focus on the density (matrix) instead of the energy
(scalar) is beneficial since it provides the directional tensor description
of the polarization that could be combined with other effects such as Pauli repulsion or dispersion,
in contrast to the standard interaction energy approaches. We show that accurate polarization 
of the one\hyp{}particle density
matrix is achievable and can be realized as long as the one\hyp{}particle density matrices
are available. It is believed that the proposed method
could be a basis for a construction of an 
alternative approach in fragment\hyp{}based
modelling of extended molecular aggregates that is not limitted to the ground state chemistry.
We introduce the necessary theory and derive the \emph{ab initio} density matrix susceptibility model
at the HF approximation in Section II. This model serves as a basis for generalization to the 
arbitrary one\hyp{}electron density with electronic correlation which is derived in Section II. 
The implementation details in a computer code as well as the validation procedures are detailed
in Section III. We then discuss the performance of the models in Section IV based on a few small molecules
in varying electric field distributions. Finally, in Section V we conclude the presentation with a few 
remarks for future development.

@@

In the Hartree\hyp{}Fock theory, ${\bf D}$ is approximated as
%
\begin{equation}
 {\bf D} = {\bf C}{\bf C}^\dagger \;,
\end{equation}
%
where ${\bf C}$ describes the occupied molecular orbitals as linear combinations
of atomic orbitals (AO's), referred here to as the LCAO-MO matrix.
${\bf D}$ is an idempotent matrix and if expressed in orthogonal basis the following holds
%
\begin{equation}
 {\bf D}^2 = {\bf D} \;.
\end{equation}
%
The change in the molecular orbitals can be parameterized as suggested by McWeeny\cite{McWeeny.RevModPhys.1960}
by defining the idempotency\hyp{}preserving variation
%
\begin{equation}
 \delta {\bf C} = {\BM\Delta} \cdot {\bf C}^{(0)} \;,
\end{equation}
%
where $\BM\Delta$ is a square non\hyp{}singular and non\hyp{}symmetric matrix.
McWeeny showed that
%
\begin{equation} \label{e:dmat-change.exact}
 \delta {\bf D} = -{\bf D}^{(0)} + \left[ {\bf D}^{(0)} + {\bf v} \right]
                                   \left[ {\bf 1} + {\bf v}^\dagger{\bf v} \right]^{-1}
                                   \left[ {\bf D}^{(0)} + {\bf v}^\dagger \right] \;,
\end{equation}
%
where the auxiliary vector ${\bf v}$ is defined as
%
\begin{equation}
 {\bf v} \equiv \left[ {\bf 1} - {\bf D}^{(0)} \right] {\BM\Delta} {\bf D}^{(0)}  \;.
\end{equation}
%
Expanding Eq.~\eqref{e:dmat-change.exact} in a Taylor series around ${\bf v}={\bf 0}$ and
truncating it at linear terms with respect to ${\BM\Delta}$ gives
%
\begin{subequations} 
 \begin{align}
 \delta {\bf D} &\cong \left[ {\bf 1} - {\bf D}^{(0)} \right] {\BM\Delta} {\bf D}^{(0)} + 
                        {\bf D}^{(0)} {\BM\Delta}^\dagger \left[ {\bf 1} - {\bf D}^{(0)} \right]  
 \label{e:dD-1} \\  &= 
  \delta {\bf C}  {\bf C}^{(0)\dagger} + {\bf C}^{(0)}\delta {\bf C}^\dagger
           - {\bf D}^{(0)} \delta {\bf C}  {\bf C}^{{(0)}\dagger} - {\bf C}^{(0)}\delta {\bf C}^\dagger {\bf D}^{(0)} 
 \label{e:dD-2}
 \end{align}
\end{subequations}
%
The latter equality is true only if the basis set functions are orthogonal, i.e., 
${\bf C}^{(0)\dagger}{\bf C}^{(0)}={\bf 1}$,
which will be the case throughout the entire work, unless stated differently.
Substituting the result from Eq.~\eqref{e:dD-1}
into Eq.~\eqref{e:dmu-exact} leads to
%%
%\begin{equation} \label{e:delta-Taylor}
% \delta {\bf D} = \left[ {\bf v} + {\bf v}^\dagger \right]
%                + \left[ {\bf v}{\bf v}^\dagger - {\bf v}^\dagger{\bf v}\right] + \ldots 
%\end{equation}
%
\begin{multline} \label{e:dmu-4-exact.linear-approximation}
 \frac{1}{2} 
 \delta {\BM{\upmu}}
  \cong
   \Tr{ 
    \left[ 
         {\mathbb{M}} {\BM\Delta} {\bf D}^{(0)}  
    \right] }
%
  +\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta}^{\dagger}
    \right] } \\
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta} {\bf D}^{(0)}
    \right] }
%
  -\Tr{ 
    \left[ 
         {\mathbb{M}} {\bf D}^{(0)} {\BM\Delta}^{\dagger} {\bf D}^{(0)}
    \right] } \;,
\end{multline}
%
It can be proven (see Appendix~\ref{a:orig-dep}) that such obtained
induced dipole moment is independent on the origin. 
This quarantees that the theory fulfills the total charge conservation requirement.
Therefore,
in the above result and also in later course of the Paper the dipole origin will be omitted. 
%where the dipole origin was omitted for notational clarity.

@@

      = \left[  {\bf C}^\dagger \mathbb{M} \left( \bf{1} - \bf{D}^{(0)} \right) \right]_{i\alpha}

@@

The theory that is necessary to validate the density matrix susceptibility models from Section~\ref{s:2}
was implemented in our in\hyp{}house plugin to {\sc Psi4} quantum chemistry program.\cite{Psi4.JCTC.2017}
Water molecule, energy\hyp{}optimized by using the HF/6-311++G** 
method\cite{Roothaan.RevModPhys.1951,Krishnan.Binkley.Seeger.Pople.JCP.1980}, 
was selected as a test system which was subject to 
electrostatic perturbations due to random collections of point charges
generating non\hyp{}uniform electric fields. 
Cartesian coordinates of perturbing charges were drawn from uniform distribution of a set of fourty
three\hyp{}dimensional Euclidean points within a sphere of radius 6.0 Bohrs
and origin in the center of mass of water molecule.
In selection of random points, space inside the union of spheres surrounding
oxygen and hydrogen atoms with radii of 4.5 and 4.0 Bohrs, respectively, was
kept free from any point charges due to potential problems with the convergence of the 
self\hyp{}consistent field (SCF) calculations.
Charge values were selected from the one\hyp{}dimensional uniform distribution
of double precision numbers in the interval of $\vert q_{\rm max} \vert$, where $q_{\rm max}$
was tuned to obtain a desired ensemble of weak, moderate and strong electric field magnitudes
(with $q_{\rm max}=0.05$, $0.2$ and $0.4$ a.u., accordingly; cf. Table~\ref{t:fields}).

%For finding a general unitary transformations,
%100 random point charge distributions were selected for which an exact density matrices
%were calculated.
In order to validate all the models discussed in this Contribution,
the electric field\hyp{}induced interaction energy, 
the dipole moment and the quadrupole moment were analysed in detail.
The electric field\hyp{}induced interaction energy at the HF level, $\delta E$,
was computed from the following exact formula:
%
\begin{multline}\label{e:dE}
 \delta E[{\bf F}] = E_{\rm nuc}({\bf F}) +
                     \Tr{\left\{ \delta {\bf D} \cdot
                                \left( {\bf H}^{(0)} + {\bf f}^{(0)} \right) \right\}} \\
                   + \Tr{\left\{ \left( {\bf D}^{(0)} + \delta {\bf D} \right) \cdot
                                \left( 2{\bf V} + 2\delta {\bf J} - \delta {\bf K}\right) \right\}} \;.
\end{multline}
%
In Eq.~\eqref{e:dE}, ${\bf f}^{(0)}$ and ${\bf H}^{(0)}$ are the Fock matrix and 
the one\hyp{}electron Hamiltonian matrix of an unperturbed HF ground state,
${\bf V}$ is the one\hyp{}electron external electrostatic potential matrix whereas $\delta {\bf J}$
and $\delta {\bf K}$ are electric field\hyp{}induced changes in the Coulomb and the exchange HF matrices due to 
the polarization of the electron density distribution, respectively. $E_{\rm nuc}({\bf F})$ denotes the interaction energy
due to the nuclei.
In order to compare the performance of the distributed polarizability approach,
the approximate interaction energy was also evaluated by using the expression:
%
\begin{multline}\label{e:dE-dpol}
 \delta E[{\bf F}] \approx E_{\rm nuc}({\bf F}) + 
                     2 \Tr{\left\{ {\bf D}^{(0)} {\bf V}\right\} }  \\
                   - \frac{1}{2} \sum_i^{\rm occ} {\BM \alpha}_i : {\bf F}({\bf r}_i) \otimes {\bf F}({\bf r}_i) \;,
\end{multline}
%
in which ${\BM \alpha}_i$ is the distributed anisotropic polarizability tensor associated with the
$i$th localized molecular orbital (LMO) centroid (see Eq.~\eqref{e:dmu-dpol}). 
%, ${\bf r}_i = \tbraket{\phi_i}{\bf r}{\phi_i}$.
The distributed polarizabilities were computed by utilizing the coupled\hyp{}perturbed Hartree\hyp{}Fock
method.\cite{McWeeny.RevModPhys.1960,Dodds.McWeeney.Sadlej.MolPhys.1977} 
Orbital localization was performed by using the Boys method.\cite{Boys.RevModPhys.1960} 
Note that in Eq.~\eqref{e:dE-dpol}, 
multipole approximation is introduced only for the part associated with the electric field\hyp{}induced
change in the one\hyp{}particle density matrix, i.e., the polarization of the electron density distribution.
%This is because we focus in this Work on the pure polarization process. Therefore, comparison between
%the results obtained by using Eq.~\eqref{e:dE-dpol} and Eq.~\eqref{e:dE} will reveal

The electric field\hyp{}induced dipole and primitive quadrupole moments were evaluated from the following expressions:
%
\begin{subequations}\label{e:dmult}
  \begin{align}
   \delta \upmu_{u}   &= 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \tbraket{\alpha}{u}{\beta}  \;,\\
   \delta \Theta_{uw} &= 2\sum_{\alpha\beta} \delta D_{\alpha\beta} \tbraket{\alpha}{(u-u_{\rm CM})(w-w_{\rm CM})}{\beta} \;. 
  \end{align}
\end{subequations}
%
where the subscript CM denotes the center of mass of water molecule.
When the distributed polarizabilities were used instead of the polarization\hyp{}induced 
difference density matrices, primitive induced quadrupole moments were computed from
the induced dipole moments according to
%
\begin{equation}
  \delta {\BM \Theta} = \sum_i^{\rm occ} 
   \left\{ 
       \delta {\BM \upmu}_i \otimes \left( {\bf r}_i - {\bf r}_{\rm CM} \right) 
      + \left( {\bf r}_i - {\bf r}_{\rm CM} \right) \otimes \delta {\BM \upmu}_i
   \right\} \;.
\end{equation}
%
%where ${\bf r}_{\rm CM}$ is the center of mass of water molecule.
Primitive quadrupole moments were subsequently transformed to their traceless forms by
%
\begin{equation}
  \delta \widetilde{\Theta}_{uw} = \delta \Theta_{uw} - \frac{1}{3} \delta_{uw} \Tr{\BM\Theta} \;.
\end{equation}
%
Methodology for the calculation of the electric field and electric field gradient due to point charges
can be found elsewhere.\cite{Blasiak.Cho.JCP.2014}
